# Hypotheses and associations {#sec-hypotheses-associations}

::: callout-warning
Under construction
:::

```{r libraries-hide}
#| warning: false
#| echo: false
library(tidyverse)
```

```{r readin-hide}
#| warning: false
#| echo: false
# -- code to read in the aggregated subject level data from the clearly-understood data set
# -- study one
study.one.gen <- read_csv("study-one-general-participants.csv", 
                                  col_types = cols(
                                    participant_ID = col_factor(),
                                    study = col_factor(),
                                    EDUCATION = col_factor(),
                                    GENDER = col_factor(),
                                    ETHNICITY = col_factor()
                                    )
                                   )

# -- study two
study.two.gen <- read_csv("study-two-general-participants.csv", 
                                  col_types = cols(
                                    participant_ID = col_factor(),
                                    study = col_factor(),
                                    EDUCATION = col_factor(),
                                    GENDER = col_factor(),
                                    ETHNICITY = col_factor()
                                    )
                                   )
```

## Overview {#sec-associations-intro} 

Welcome to your overview of the materials you will work with in **PSYC401 Week 6**.
We will complete five classes in weeks 6-10.
These classes are designed to help students to learn, consolidate and revise some core psychological data analysis methods.
We will do this in the context of a live research project with potential real world impacts: the **Clearly understood** project.

Our focus will be on **what makes it easy or difficult for people to understand written health information**.
We encounter written health information all the time: in warnings signs, on medication labels, in clinics when we go to see the doctor, and online when we research things we are worried about.
It is not always easy to understand this information.
The problem is that it is unclear how health information should be communicated.

1. What person attributes predict success in understanding?
2. Can people accurately evaluate whether they correctly understand written health information?

We will present our PSYC401 lessons in the context of this research project because we think that this *context* will help you to make sense of the data, and to see why we ask you to practice the skills we are teaching.

We will be revisiting some of the ideas and techniques you have seen introduced in previous classes.
This is to give you the opportunity to consolidate your learning.
We will be extending your development with some new ideas, to strengthen your skills.

## Our learning goals {#sec-associations-learning-goals} 

### Learning targets for this week {#sec-associations-learning-goals-week6} 

1. Critical thinking: hypotheses

-   *Concepts*: begin with critical thinking
-   *Skills*: developing hypotheses

2. Data analysis: associations

-   *Concepts* -- associations: correlations, estimates and hypothesis tests
-   *Skills* -- visualizing variation and covariation
-   *Skills* -- writing the code
-   *Skills* -- estimating correlations
-   *Skills* -- hypothesis tests for correlations
-   *Skills* -- interpreting and reporting correlations

## Resources for you {#sec-associations-resources} 

You will see -- below -- links to the lectures, information about the data we will analyze, and an explanation of the activities.

All the links to the lectures, and everything you need for your practical work class can *also* be found in the **Week 6 files** folder on Moodle [here](https://modules.lancaster.ac.uk/course/view.php?id=40622#coursecontentcollapse7){target="_blank"}

### Lectures: video recordings

The lecture material for this week is presented in five short parts.
Click on a link and your browser should open a tab showing the *Panopto* video for the lecture part.

[Part 1 of 5](https://lancaster.cloud.panopto.eu/Panopto/Pages/Viewer.aspx?id=3e5f2ca1-fc6d-4e31-8f7c-af5000c9916f){target="_blank"}

[Part 2 of 5](https://lancaster.cloud.panopto.eu/Panopto/Pages/Viewer.aspx?id=9a10462c-e185-4069-b72a-af5000cc076a){target="_blank"}

[Part 3 of 5](https://lancaster.cloud.panopto.eu/Panopto/Pages/Viewer.aspx?id=1653a285-fa5b-45c4-a850-af5000d1c8d5){target="_blank"}

[Part 4 of 5](https://lancaster.cloud.panopto.eu/Panopto/Pages/Viewer.aspx?id=ca174423-d385-40d5-9068-af5000d60688){target="_blank"}

[Part 5 of 5](https://lancaster.cloud.panopto.eu/Panopto/Pages/Viewer.aspx?id=9d53628e-3133-43aa-b78f-af5000dc0d0b){target="_blank"}

<!-- ## Pre-lab activity 1 -->

<!-- In weeks 16-19, we will be working together on a research project to investigate how people vary in their response to health advice. -->
<!-- Completing the project involves collecting responses from PSYC122 students. -->
<!-- In our class activities, we can then analyze the data we collect. -->

<!-- To enter your responses, we invite you to complete a short survey. -->
<!-- You can complete the survey here (just click on the web address to get started): -->

<!-- [Complete the survey](https://lancasteruni.eu.qualtrics.com/jfe/form/SV_0ww8HHxx172TPJs){target="_blank"} -->

<!-- ### Survey information -->

<!-- The survey asks you to: -->

<!-- - complete some questions about who you are; -->
<!-- - and then answer some questions about what you know about some English words, about what you know about health matters, and about how you approach reading. -->

<!-- The survey then asks you to: -->

<!-- - read five short extracts from patient information leaflets about different kinds of health issue; -->
<!-- - respond to some multiple choice questions about each extract; -->
<!-- - and rate how well you think you understand the advice. -->

<!-- The survey should take about 20 minutes to complete. -->
<!-- Some people will take less time, and some people might take a little more time. -->

<!-- Taking part in the survey is **completely voluntary**. -->
<!-- You can stop at any time without completing the survey if you do not want to finish it. -->
<!-- If you do not want to do the survey, you can do an alternative activity (see below). -->

<!-- All responses will be recorded completely anonymously. -->

<!-- ## Pre-lab activity 1 alternative -->

<!-- If you do not want to complete the survey, we invite you to read the pre-registered research plan for the *PSYC122 health advice* research project. -->

<!-- [Read the project pre-registration](https://osf.io/p6fsc/){target="_blank"} -->

## Pre-lab activity 2: Getting ready for the lab class

### Get your files ready

Download the [PSYC401_week6_materials.zip](files/PSYC401_week6_materials.zip) files you need and upload them to your RStudio Server.

The folder includes data files:

- `study-one-general-participants.csv`
- `study-two-general-participants.csv`

and code files:

- `401-associations-how-to.R`
- `401-associations-workbook.R`

You will use these files for your practical learning.

### Review the how-to guide

We show you how to do everything you need to do in the lab activity (see the next section) in the `how-to` guide.

The guide comprises an .R file `401-associations-how-to.R` with code and advice.
The code in the .R file was written to work with the data file `study-one-general-participants.csv`.

<!-- In the video, I work my way through the R code, step-by-step, explaining what each line of code does. -->
<!-- Watching the video takes about 20 minutes. -->
<!-- It is completely optional but I have provided the video because students have told us they find it useful. -->

<!-- [Watch the how-to guide](https://dtu-panopto.lancs.ac.uk/Panopto/Pages/Viewer.aspx?id=fc290f97-5085-4b92-8870-ae3f01294230){target="_blank"} -->

### Workbook

In the workbook .R file:

- `401-associations-workbook.R`

you will work with data from a study about how people respond to guidance about a variety of health topics (general topics)

- `study-two-general-participants.csv`

If you have not downloaded them already, the workbook .R file and the data .csv file can be downloaded from:

[PSYC401_week6_materials.zip](files/PSYC401_week6_materials.zip)

Or you can download them from the module [Moodle page for PSYC401](https://modules.lancaster.ac.uk/course/view.php?id=40622#coursecontentcollapse7){target="_blank"}

#### What is in the workbook .R file?

We will take things step-by-step.
We will split .R scripts into parts, tasks and questions:  

- different parts for different phases of the analysis workflow;
- different tasks for different steps in each phase;
- different questions to examine different ideas or coding steps.

#### Tasks

We are going to work through the following tasks:

1. Empty the R environment -- using `rm(list=ls())`
2. Load relevant libraries -- using `library()`
3. Read in the data file -- using `read_csv()`
4. Inspect the data -- using `head()` and `summary()`
5. Change the type classification of a variable in the data -- using `as.factor()`
6. Draw histograms to examine the distributions of variables -- using `ggplot()` and `geom_histogram()`
7. Draw scatterplots to examine the association between pairs of variables -- using `ggplot()` and `geom_point()`
8. Estimate and test the correlations between pairs of variables -- using `cor.test()`

The activity `401-associations-workbook.R` file takes you through the tasks, one by one.

If you are unsure about what you need to do, check in the `how-to` guide: look at the advice in `401-associations-how-to.R`.

You will see that you can match a task in the `activity` to the same task in the `how-to`.
The `how-to` shows you what function you need and how you should write the function code.
You will need to change the names of the dataset or the variables to complete the tasks in the activity.

This process of adapting demonstration code is a process critical to data literacy and to effective problem solving in working with data in psychological science.

### What is in the data files

Each of the data files we will work with has a similar structure.
```{r head}
head(study.two.gen)
```

You can see the columns:

- `participant_ID` participant code
- `mean.acc` average accuracy of response to questions testing understanding of health guidance
- `mean.self` average self-rated accuracy of understanding of health guidance
- `study` varianble coding for what study the data were collected in
- `AGE` age in years
- `HLVA` health literacy test score
- `SHIPLEY` vocabulary knowledge test score
- `FACTOR3` reading strategy survey score
- `GENDER` gender code
- `EDUCATION` education level code
- `ETHNICITY` ethnicity (Office National Statistics categories) code

<!-- ### Answers -->

<!-- You can download the answers version of the activity .R file [401-associations-workbook-answers.R](files/401-associations-workbook-answers.R) -->

<!-- The answers version will present my answers for questions, and some extra information where that is helpful. -->
















