# Introduction to visualization {#sec-vis-intro-intro}

```{r library}
#| echo: false
#| warning: false
library(datasauRus)
library(faux)
library(ggExtra)
library(kableExtra)
library(knitr)
library(lme4)
library(patchwork)
library(tidyverse)
```

```{r readin}
#| echo: false
#| warning: false

# -- read in the datasaurus dataset:
datasaurus <- read_csv("Datasaurus_data.csv")

# -- code to read in the aggregated subject level data from the clearly-understood data set
# -- study one
clearly.one.subjects <- read_csv("study-one-general-participants.csv", 
                                  col_types = cols(
                                    participant_ID = col_factor(),
                                    study = col_factor(),
                                    EDUCATION = col_factor(),
                                    GENDER = col_factor(),
                                    ETHNICITY = col_factor()
                                    )
                                   )

# -- code to read in the aggregated subject level data from the clearly-understood data set
# -- study two -- general texts
clearly.two.subjects <- read_csv("study-two-general-participants.csv", 
                                 col_types = cols(
                                   participant_ID = col_factor(),
                                   study = col_factor(),
                                   EDUCATION = col_factor(),
                                   GENDER = col_factor(),
                                   ETHNICITY = col_factor()
                                   )
                                  )

```

## Overview {#sec-vis-intro-overview}

Welcome to your overview of the work we will do together in **Week 9**.

This week, we will focus on perspectives and practices in data visualization.

In a wide range of professions, you will benefit if you can exercise skills in the capacity to produce and to interpret data visualizations.
The power to produce informative eye-catching visualizations in R is industry-leading, everywhere from news media to medical science.
This means that learning how to produce visualizations using R will give you an advantage in many settings.

Our materials are designed to help you to think about what you are doing --- to understand the aims of the practical steps --- as well as to learn about producing professional effective data visualizations.
This is because, as you progress in your careers, you may come to manage as well as to direct the development of visualizations, so we need to **see what best practice looks like**.

We will continue to work with data collected for the *Clearly understood* project (@sec-associations) because we think that working with these data in this research **context** will help you to make sense of the data, and to see why we ask you to practise the skills we are teaching.

## Our learning goals {#sec-vis-intro-goals}

This week, we focus on *both* developing your critical thinking and strengthening your practical skills: our objectives blend *concepts* and *skills* targets.

**Our learning objectives: --- what are we learning about?**

We are working together to help you:

1.  *Goals* --- Formulate questions you can ask yourself to help you to work effectively
2.  *Audience* --- Understand the psychological factors that affect your impact
3.  *Development* --- Work reflectively through a development process
4.  *Implement* --- Produce visualizations in line with best practice

**Our assessment targets: --- how do you know if you have learned?**

We are working together so you can:

1.  *Goals* --- Identify a set of targets for a development process in your professional teams
2.  *Audience* --- Explain what you need to do to make a visualization effective
3.  *Development* --- Locate yourself within the stages of the development process
4.  *Implement* --- Produce visualizations that look good and are useful

As we work together, you will have the opportunity to consolidate your learning in coding plots in R but you can be sure, too, that the practical exercises will help you to grow your skills with new ideas.

## Learning resources {#sec-vis-intro-resources}

You will see *next* links to the lectures we created to explain the concepts we want you to learn about and the practical visualization skills we want you to develop (@sec-vis-intro-lectures), *then* information about the practical materials we have provided to help you to practise your skills (@sec-vis-intro-practical).

All the links to the lecture videos, the lecture slides, and everything you need for your practical work can *also* be found in the Week 8 files folder on Moodle [here](https://modules.lancaster.ac.uk/course/view.php?id=40622#coursecontentcollapse9){target="_blank"}

In @sec-vis-intro-notes, we present the lecture slide points. We do this here because we can share the code we used to generate the plots we use in some of the slides [^visualization-intro-1]

[^visualization-intro-1]: We write the slides and this book in [Quarto](https://quarto.org){target="_blank"} in R-Studio. Quarto scripts can be rendered as .html to share web-books like this one, or to share slides like those we use in presenting the lecture. One advantage of using Quarto is that we can share a plot and the code we used to generate the plot in the same page.

::: callout-tip
**Linked resources** include:

-   In @sec-visualization, we present a more extensive discussion of data visualization, elaborating on some ideas, incorporating additional example plots and enabling you to work with alternate data-sets. This is provided as *optional* reading and may support more advanced development for students interested in future professional roles involving data analysis or data visualization.
-   In the *PSYC403* Week 8, I present a lecture that outlines some *perspectives* or useful ways of thinking about data visualization: the history; the context in professional work; and what research suggests are effective ways to produce visualizations. You can access the resources for that class [here](https://modules.lancaster.ac.uk/course/view.php?id=40624#coursecontentcollapse9){target="_blank"}.
:::

### Lectures {#sec-vis-intro-lectures}

The lecture material for this week is presented in four short parts. Click on a link and your browser should open a tab showing the *Panopto* video for the lecture part.

[Part 1 of 4](https://lancaster.cloud.panopto.eu/Panopto/Pages/Viewer.aspx?id=b7324096-ad13-4ca1-b8ab-b0c000d85850){target="_blank"}

[Part 2 of 4](https://lancaster.cloud.panopto.eu/Panopto/Pages/Viewer.aspx?id=8f42d056-6da6-4aba-b893-b0c000dc28d1){target="_blank"}

[Part 3 of 4](https://lancaster.cloud.panopto.eu/Panopto/Pages/Viewer.aspx?id=05e049cb-277d-4d76-888d-b0c000df3c6f){target="_blank"}

[Part 4 of 4](https://lancaster.cloud.panopto.eu/Panopto/Pages/Viewer.aspx?id=5fccaa55-9d07-4e24-8995-b0c0010fe92b){target="_blank"}

You can download the slides we presented in the lecture in two different formats, depending on what you think will be most useful to you:

-   Download the slides exactly as they appear in the lecture from [this link](files/401-data-visualisation-practices.html). The .html file can be opened and viewed in any web browser (e.g., Chrome, Firefox, Safari).
-   Or you can download a printable Word .docx presentation of the slides from [this link](files/401-data-visualisation-practices-printable.docx). The .docx can be opened in Microsoft Word. The figures will not appear exactly as they do in the lecture recording because Word cannot cope with images so well but the trade-off is that you get a document you can print and edit to add notes.

### Practical materials {#sec-vis-intro-practical}

We have collected the practical materials together into a folder.

The folder includes the data files:

-   `study-one-general-participants.csv`
-   `study-two-general-participants.csv`

and .R code files:

-   `401-visualization-how-to.R`
-   `401-visualization-workbook.R`

You will use these files for your practical learning.

You can download the .R files and the data .csv files in a single folder, using the link [here](files/PSYC401_visualization_materials.zip).

Or you can download the files as individual files from the module [Moodle page for PSYC401](https://modules.lancaster.ac.uk/course/view.php?id=40622#coursecontentcollapse9){target="_blank"}.

Once you have downloaded the file folder, you will need to upload it to the R-Studio server to access and use the R files.

#### The how-to guide {#sec-vis-intro-practical-how-to}

In the `how-to` guide:

-   `401-visualization-how-to.R`

we show you how to do everything you need to do in the practical workbook (see @sec-vis-intro-practical-workbook). The guide comprises an .R file `401-visualization-how-to.R` with code and advice.

The code in the .R file was written to work with the data file

-   `study-one-general-participants.csv`.

::: callout-tip
-   Work through the steps in the `how-to` guide first, this practice will help you to understand what you need to do for the `workbook` tasks.

-   The `how-to` guide and the `workbook` have similar structures. This is intentional: so that you can copy and *adapt* code from the `how-to` guide to do the practical tasks in the `workbook`.
:::

### The workbook {#sec-vis-intro-practical-workbook}

In the `workbook`:

-   `401-visualization-workbook.R`

you will work with the data file

-   `study-two-general-participants.csv`

We split .R scripts into parts, tasks and questions.

For this class on data visualization practices, our practical materials have two aims:

1.  Helping you to learn about data visualization;
2.  Helping you to learn how to help yourself by accessing, evaluating and exploiting the rich R knowledge ecosystem.

This means that in both the `how-to` and the `workbook` the different parts concern different sources of online information and how to access each.

In the `how-to` and the `workbook` we look at methods to produce visualizations that display information about both summary statistics (e.g., the average outcome) and raw or individual outcome variability. We explain why we should visualize data like this in the lecture (as you can see in @sec-vis-intro-lectures or @sec-vis-intro-notes).

Specifically, the materials are written to support learning how to work with visualizations called *boxplots* and *rain cloud* plots. These are popular data visualization techniques so it is important to learn how to build them and how to interpret them. If you have already been introduced to them, you will find that our materials here show how to edit or polish the visualizations to make them more effective.

There are three main sources of information you can access for free online.

1.  The people who write software like the `{tidyverse}` or `{ggplot2}` libraries provide manuals, reference guides and tutorials. This information is often written as free web books, or as hard copy books.

2.  These people, and often ther people, may write tutorials or guides or teaching materials designed to show learners (like us) how to use R functions or do certain things using R. They may present these tutorials or guides as web books, blog sites or video tutorials e.g. on Youtube or TikTok.

3.  Many post questions and answers to discussion forums like *Stackoverflow*.

Learning how to find, understand and use this information teaches two lessons:

-   A lot of scholarly and technical information is online and free.
-   Learning how to access this information is a key way that most professionals work out what they want to do and how they can do it.

In our practical materials, we look at how you can:

-   search for and find relevant information;
-   learn how to understand the tools using the information shared through online sources;
-   work with example or demonstration code, adapting it for your own purposes.

::: callout-tip
This process of adapting demonstration code is a process critical to data literacy and to effective problem solving in working with data in psychological science.
:::

**Part 3** in the `workbook` focuses on locating and using `{ggplot2}` reference information to learn how to build box plots.

**Part 4** in the `workbook` focuses on locating and using online tutorial information or how-to guides to build *rain cloud* plots. This kind of plot incorporates elements of the boxplot, the scatterplot and the density plot to give the viewer summary information about the distribution of scores on a variable while also giving information about the variability of outcomes.

**Part 5** in the `workbook` focuses on using information available in the public discussion *Stackoverflow* so that you can learn how to export the plots you make, in order to include them in reports.

The activity `401-visualization-workbook.R` file takes you through the tasks, one by one.

If you are unsure about what you need to do, check the advice in `401-visualization-how-to.R`.

You will see that you can match a task in the `activity` to the same task in the `how-to`. The `how-to` shows you what function you need and how you should write the function code. You will need to change the names of the data-set or the variables to complete the tasks in the activity.

### The data files

Each of the data files we will work with has a similar structure.

Here are what the first few rows in the data file `study-two-general-participants.csv` looks like:

```{r head}
#| warning: false
#| echo: false
kable(head(clearly.two.subjects))
```

You can see the columns:

-   `participant_ID` participant code
-   `mean.acc` average accuracy of response to questions testing understanding of health guidance
-   `mean.self` average self-rated accuracy of understanding of health guidance
-   `study` variable coding for what study the data were collected in
-   `AGE` age in years
-   `HLVA` health literacy test score
-   `SHIPLEY` vocabulary knowledge test score
-   `FACTOR3` reading strategy survey score
-   `GENDER` gender code
-   `EDUCATION` education level code
-   `ETHNICITY` ethnicity (Office National Statistics categories) code

### The answers

After the practical class, you will be able to download the answers version of the `workbook` here.

<!-- [401-visualization-workbook-answers.R](files/401-visualization-workbook-answers.R) -->

The answers version will present my answers for questions, and some extra information where that is helpful.







### Summary

You start your work with these questions:

1.  What are our goals?
2.  What does our audience need or expect?

You develop your visualization in a reflective process:

1.  Begin with a quick draft to show the distributions or make the comparisons you think about first
2.  Then reflect, and edit: does this enable me to discover sources of variability in my data?
3.  Then reflect, and edit: does this enable me to effectively communicate what I want to communicate?
4.  Then reflect, and edit: does this look good? -- do my viewers tell me this works well?

::: callout-tip
I can only show you the potential for creative and effective visualization

-   experiment and find what looks good and is useful to you
-   seek out information -- good places to start are:

<https://ggplot2.tidyverse.org/index.html>

<https://r-graph-gallery.com>
:::
