# Introduction to visualization {#sec-vis-intro-intro}

```{r library}
#| echo: false
#| warning: false
library(datasauRus)
library(faux)
library(ggExtra)
library(kableExtra)
library(knitr)
library(lme4)
library(patchwork)
library(tidyverse)
```

```{r readin}
#| echo: false
#| warning: false

# -- read in the datasaurus dataset:
datasaurus <- read_csv("Datasaurus_data.csv")

# -- code to read in the aggregated subject level data from the clearly-understood data set
# -- study one
clearly.one.subjects <- read_csv("study-one-general-participants.csv", 
                                  col_types = cols(
                                    participant_ID = col_factor(),
                                    study = col_factor(),
                                    EDUCATION = col_factor(),
                                    GENDER = col_factor(),
                                    ETHNICITY = col_factor()
                                    )
                                   )

# -- code to read in the aggregated subject level data from the clearly-understood data set
# -- study two -- general texts
clearly.two.subjects <- read_csv("study-two-general-participants.csv", 
                                 col_types = cols(
                                   participant_ID = col_factor(),
                                   study = col_factor(),
                                   EDUCATION = col_factor(),
                                   GENDER = col_factor(),
                                   ETHNICITY = col_factor()
                                   )
                                  )

```

## Overview {#sec-vis-intro-overview}

Welcome to your overview of the work we will do together in **Week 9**.

This week, we will focus on perspectives and practices in data visualization.

In a wide range of professions, you will benefit if you can exercise skills in the capacity to produce and to interpret data visualizations.
The power to produce informative eye-catching visualizations in R is industry-leading, everywhere from news media to medical science.
This means that learning how to produce visualizations using R will give you an advantage in many settings.

Our materials are designed to help you to think about what you are doing --- to understand the aims of the practical steps --- as well as to learn about producing professional effective data visualizations.
This is because, as you progress in your careers, you may come to manage as well as to direct the development of visualizations, so we need to **see what best practice looks like**.

We will continue to work with data collected for the *Clearly understood* project (@sec-associations) because we think that working with these data in this research **context** will help you to make sense of the data, and to see why we ask you to practise the skills we are teaching.

## Our learning goals {#sec-vis-intro-goals}

This week, we focus on *both* developing your critical thinking and strengthening your practical skills: our objectives blend *concepts* and *skills* targets.

**Our learning objectives: --- what are we learning about?**

We are working together to help you:

1.  *Goals* --- Formulate questions you can ask yourself to help you to work effectively
2.  *Audience* --- Understand the psychological factors that affect your impact
3.  *Development* --- Work reflectively through a development process
4.  *Implement* --- Produce visualizations in line with best practice

**Our assessment targets: --- how do you know if you have learned?**

We are working together so you can:

1.  *Goals* --- Identify a set of targets for a development process in your professional teams
2.  *Audience* --- Explain what you need to do to make a visualization effective
3.  *Development* --- Locate yourself within the stages of the development process
4.  *Implement* --- Produce visualizations that look good and are useful

As we work together, you will have the opportunity to consolidate your learning in coding plots in R but you can be sure, too, that the practical exercises will help you to grow your skills with new ideas.

## Learning resources {#sec-vis-intro-resources}








### Summary

You start your work with these questions:

1.  What are our goals?
2.  What does our audience need or expect?

You develop your visualization in a reflective process:

1.  Begin with a quick draft to show the distributions or make the comparisons you think about first
2.  Then reflect, and edit: does this enable me to discover sources of variability in my data?
3.  Then reflect, and edit: does this enable me to effectively communicate what I want to communicate?
4.  Then reflect, and edit: does this look good? -- do my viewers tell me this works well?

::: callout-tip
I can only show you the potential for creative and effective visualization

-   experiment and find what looks good and is useful to you
-   seek out information -- good places to start are:

<https://ggplot2.tidyverse.org/index.html>

<https://r-graph-gallery.com>
:::
